---
title: "isoTensor: Cell-type-resolved isoform abundance estimation from bulk RNA-seq"
author:
- name: Koki Tsuyuzaki
  email: k.t.the-answer@hotmail.co.jp
date: "`r Sys.Date()`"
package: isoTensor
output: BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{isoTensor: Cell-type-resolved isoform abundance estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`isoTensor` estimates cell-type-resolved isoform abundance from bulk RNA-seq.
It takes two inputs:

- $X \in \mathbb{R}_+^{N \times I}$: a bulk isoform abundance matrix
  (individual $\times$ isoform), and
- $A \in \mathbb{R}_+^{N \times C}$: a cell type fraction matrix
  (individual $\times$ cell type), estimated in advance by an external
  reference-based deconvolution tool such as Bisque.

Estimating the cell type fractions themselves is **not** the responsibility of
`isoTensor`; $A$ is always treated as fixed. `isoTensor` estimates the
population-level cell-type-by-isoform abundance matrix
$B \in \mathbb{R}_+^{C \times I}$ under the model

$$
X \approx A B,
\qquad
\min_{B \geq 0} \frac{1}{2} \| X - A B \|_F^2,
$$

by the multiplicative update rule

$$
B \leftarrow B \odot
\frac{A^\top X}{A^\top A B + \varepsilon}.
$$

# Toy data

A small synthetic dataset with known ground truth is available from
`isoToyModel`:

```{r data, echo=TRUE}
library("isoTensor")
toy <- isoToyModel(N=5, C=2, I=6, seed=1234)
str(toy)
```

Here `toy$X` is exactly `toy$A %*% toy$B`.

# Fitting the matrix model

```{r fit, echo=TRUE}
set.seed(1234)
fit <- isoTensor(toy$X, toy$A, model="matrix", num.iter=1000, thr=1e-12)
```

With noiseless data, the ground truth `B` is recovered almost exactly:

```{r recovery, echo=TRUE}
max(abs(fit$B - toy$B))
```

The reconstruction error monotonically decreases:

```{r objective, echo=TRUE, fig.height=4, fig.width=4}
plot(as.numeric(fit$RecError[-1]), type="l", log="y",
    xlab="Iteration", ylab="||X - AB||_F")
```

# Noisy data

```{r noisy, echo=TRUE}
toy_noisy <- isoToyModel(N=5, C=2, I=6, noise=0.05, seed=42)
set.seed(42)
fit_noisy <- isoTensor(toy_noisy$X, toy_noisy$A, num.iter=1000, thr=1e-12)
norm(fit_noisy$B - toy_noisy$B, "F") / norm(toy_noisy$B, "F")
```

# The individual-specific tensor model

With `model="tensor"`, `isoTensor` estimates an individual-specific
latent tensor $B \in \mathbb{R}_+^{N \times C \times I}$ under the
partial non-negative Tucker structure

$$
B_{nci} = \sum_{p,q} U_{np}\, G_{pcq}\, W_{iq},
\qquad
X_{ni} \approx \sum_c A_{nc} B_{nci},
$$

where the cell type mode is not compressed and $A$ stays fixed. The
ranks of the individual mode ($R_N$ = `rank_individual`) and the
isoform mode ($R_I$ = `rank_isoform`) must be specified; the regime
$R_N \geq R_I$ is generically non-identifiable for $B$ and is flagged
by a warning (see `notes/tensor-model.md` in the GitHub repository).

Three optimizers minimize the same Frobenius objective. The default,
`optimizer="lbfgs"`, removes the non-negativity constraints by the log
reparameterization $U = \exp(\Theta_U)$ (and likewise for $G$, $W$) and
runs L-BFGS with analytical gradients; in synthetic benchmarks it is
several times faster than the multiplicative updates and recovers the
latent tensor substantially more accurately. `optimizer="mu"` is the
multiplicative-update baseline with a provably non-increasing
objective, and `optimizer="hybrid"` refines the MU solution with
L-BFGS.

```{r tensor, echo=TRUE}
tt <- isoToyModel(model="tensor", N=20, C=2, I=20,
    rank_individual=1, rank_isoform=2, A_variation="high", seed=10)
set.seed(10)
fit_t <- isoTensor(tt$X, tt$A, model="tensor",
    rank_individual=1, rank_isoform=2, optimizer="lbfgs")
# relative recovery error of the latent tensor
sqrt(sum((fit_t$B - tt$B)^2)) / sqrt(sum(tt$B^2))
fit_t$objective
fit_t$lbfgs$convergence
```

# Session Information {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
