---
title: "isoTensor: Cell-type-resolved isoform abundance estimation from bulk RNA-seq"
author:
- name: Koki Tsuyuzaki
  email: k.t.the-answer@hotmail.co.jp
date: "`r Sys.Date()`"
package: isoTensor
output: BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{isoTensor: Cell-type-resolved isoform abundance estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`isoTensor` estimates cell-type-resolved isoform abundance from bulk RNA-seq.
It takes two inputs:

- $X \in \mathbb{R}_+^{N \times I}$: a bulk isoform abundance matrix
  (individual $\times$ isoform), and
- $A \in \mathbb{R}_+^{N \times C}$: a cell type fraction matrix
  (individual $\times$ cell type), estimated in advance by an external
  reference-based deconvolution tool such as Bisque.

Estimating the cell type fractions themselves is **not** the responsibility of
`isoTensor`; $A$ is always treated as fixed. `isoTensor` estimates the
population-level cell-type-by-isoform abundance matrix
$B \in \mathbb{R}_+^{C \times I}$ under the model

$$
X \approx A B,
\qquad
\min_{B \geq 0} \frac{1}{2} \| X - A B \|_F^2,
$$

by the multiplicative update rule

$$
B \leftarrow B \odot
\frac{A^\top X}{A^\top A B + \varepsilon}.
$$

# Toy data

A small synthetic dataset with known ground truth is available from
`isoToyModel`:

```{r data, echo=TRUE}
library("isoTensor")
toy <- isoToyModel(N=5, C=2, I=6, seed=1234)
str(toy)
```

Here `toy$X` is exactly `toy$A %*% toy$B`.

# Fitting the matrix model

```{r fit, echo=TRUE}
set.seed(1234)
fit <- isoTensor(toy$X, toy$A, model="matrix", num.iter=1000, thr=1e-12)
```

With noiseless data, the ground truth `B` is recovered almost exactly:

```{r recovery, echo=TRUE}
max(abs(fit$B - toy$B))
```

The reconstruction error monotonically decreases:

```{r objective, echo=TRUE, fig.height=4, fig.width=4}
plot(as.numeric(fit$RecError[-1]), type="l", log="y",
    xlab="Iteration", ylab="||X - AB||_F")
```

# Noisy data

```{r noisy, echo=TRUE}
toy_noisy <- isoToyModel(N=5, C=2, I=6, noise=0.05, seed=42)
set.seed(42)
fit_noisy <- isoTensor(toy_noisy$X, toy_noisy$A, num.iter=1000, thr=1e-12)
norm(fit_noisy$B - toy_noisy$B, "F") / norm(toy_noisy$B, "F")
```

# Session Information {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
